Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-06-02 newpaper/paper.tex
wip

\usepackage{xcolor}
\usepackage{mathpartir}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{scalerel}
\usepackage{bm}
\usepackage{mathtools}

\newcommand{\mit}[1]{{\mathsf{#1}}}
\newcommand{\msf}[1]{{\mathsf{#1}}}
\newcommand{\mbf}[1]{{\mathbf{#1}}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\wh}[1]{\widehat{#1}}
\newcommand{\mdo}{\mbf{do}\,}
\newcommand{\ind}{\hspace{1em}}
\newcommand{\bif}{\mbf{if}\,}
\newcommand{\bthen}{\mbf{then}\,}
\newcommand{\belse}{\mbf{else}\,}
\newcommand{\return}{\mbf{return}\,}
\newcommand{\pure}{\mbf{pure}\,}
\newcommand{\lam}{\lambda\,}
\newcommand{\data}{\mbf{data}\,}
\newcommand{\where}{\mbf{where}}
\newcommand{\M}{\msf{M}}
\newcommand{\letrec}{\mbf{letrec}\,}
\newcommand{\of}{\mbf{of}\,}
\newcommand{\go}{\mit{go}}
\newcommand{\add}{\mit{add}}
\newcommand{\letdef}{\mbf{let\,}}
\newcommand{\map}{\mit{map}}

\newcommand{\vas}{\mathsf{as}}
\newcommand{\vbs}{\mathsf{bs}}
\newcommand{\vcs}{\mathsf{cs}}
\newcommand{\vxs}{\mathsf{xs}}
\newcommand{\vys}{\mathsf{ys}}
\newcommand{\vsp}{\mathsf{sp}}
\newcommand{\vma}{\mathsf{ma}}
\newcommand{\vga}{\mathsf{ga}}
\newcommand{\vm}{\mathsf{m}}
\newcommand{\vn}{\mathsf{n}}
\newcommand{\vk}{\mathsf{k}}
\newcommand{\vA}{\mathsf{A}}
\newcommand{\vB}{\mathsf{B}}
\newcommand{\vC}{\mathsf{C}}
\newcommand{\vS}{\mathsf{S}}
\newcommand{\vF}{\mathsf{F}}
\newcommand{\vR}{\mathsf{R}}
\newcommand{\vM}{\mathsf{M}}
\newcommand{\vmb}{\mathsf{mb}}
\newcommand{\mAs}{\mathsf{As}}
\newcommand{\va}{\mathsf{a}}
\newcommand{\vb}{\mathsf{b}}
\newcommand{\vc}{\mathsf{c}}
\newcommand{\vd}{\mathsf{d}}
\newcommand{\vx}{\mathsf{x}}
\newcommand{\vy}{\mathsf{y}}
\newcommand{\vz}{\mathsf{z}}
\newcommand{\vf}{\mathsf{f}}
\newcommand{\vfs}{\mathsf{fs}}
\newcommand{\vg}{\mathsf{g}}
\newcommand{\vh}{\mathsf{h}}
\newcommand{\vt}{\mathsf{t}}
\newcommand{\vs}{\mathsf{s}}
\newcommand{\vr}{\mathsf{r}}
\newcommand{\vu}{\mathsf{u}}
\newcommand{\vl}{\mathsf{l}}
\newcommand{\vns}{\mathsf{ns}}

@@

@@

\newcommand{\MTy}{\msf{MetaTy}}
\newcommand{\VTy}{\msf{ValTy}}
\newcommand{\Ty}{\msf{Ty}}
\newcommand{\CTy}{\msf{CompTy}}
\newcommand{\True}{\msf{True}}
\newcommand{\False}{\msf{False}}
\newcommand{\fst}{\msf{fst}}
\newcommand{\snd}{\msf{snd}}

@@

Modern functional programming supports many convenient abstractions. These often
come with significant runtime overheads. Sometimes the overheads are acceptable,
but in other cases compiler optimization is crucial. Monads in Haskell is an
example for the latter. Even the $\Reader$ monad, which is one of the simplest
in terms of implementation, yields large overheads when compiled without
optimizations. Consider the following:
\begin{alignat*}{3}
  &\vf :: \Int \to \Reader\,\Bool\,\Int \\
  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vx\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
\end{alignat*}
With optimizations enabled, GHC compiles this roughly to the code below:
\begin{alignat*}{3}
  &\vf :: \Int \to \Bool \to \Int \\
  &\vf = \lam \vx\,\vb.\, \bif \vb\, \bthen \vx + 10\, \belse \vx + 20
\end{alignat*}
Without optimizations we roughly get:
\begin{alignat*}{3}
  &\vf = \lam \vx.\,(\bind)\,\msf{MonadReaderDict}\,\msf{ask}\,(\lam \vb.\,\bif \vb\\
  &\ind \bthen\return\,\msf{MonadReaderDict}\,(\vx + 10)\\
  &\ind \belse\hspace{0.25em}\return\,\msf{MonadReaderDict}\,(\vx + 20))
\end{alignat*}
Here, $\msf{MonadReaderDict}$ is a runtime dictionary, containing the methods of
the $\Monad$ instance for $\Reader$, and $(\bind)\,\msf{MonadReaderDict}$ is a
field projection. Here, a runtime closure will be created for the $\lam \vb.\,...$
function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$ will create additional dynamic
closures.

The difference between optimized and unoptimized code is already large here, and
it gets even larger when we consider monad transformers or code that is
polymorphic over monads. In Haskell, such code is pervasive, even in fairly
basic programs which do not use fancy abstractions. Consider the $\msf{mapM}$
function from the Haskell Prelude:
\begin{alignat*}{3}
  & \msf{mapM} :: \Monad\,\vm \Rightarrow (\va \to \vm\,\vb) \to [\va] \to \vm\,[\vb]
\end{alignat*}
This is a third-order and second-rank polymorphic function in disguise, because
its monad dictionary argument contains the polymorphic second-order method
$(\bind)$.  Compiling $\msf{mapM}$ efficiently relies on inlining the instance
dictionary, then inlining the methods contained there, and also inlining the
functions that the higher-order binding method is applied to.

@@

\subsection{Semantics of Staging}\label{sec:semantics-of-staging}

We give a short summary of the semantics of staging here. We reuse most of the
semantics from \cite{staged2ltt} and only mention additions and differences. The
reader may refer to ibid.\ for technical details.

Staging is given by evaluation of CFTT into a presheaf model, where the base
category is given as follows: objects are typing contexts of the object language
and morphisms are parallel substitutions. The equality of morphisms is given by
strict syntactic equality of object terms. $\Ty$ is interpreted as the presheaf
which is constantly the set of object-theoretic types. Soundness of staging is
proven by a logical relation model, internally to the mentioned presheaf
model. Soundness means that the output of staging, viewed as a CFTT term, is
definitionally equal to the input.

The main addition in CFTT is the assumption of all inductive families in the
meta-level. However, based on the results in \cite{whynotw}, it is enough to
extend the meta-level syntax in \cite{staged2ltt} with W-types and identity types to
derive all inductive families. We consider the semantics of these in the following.

\subsubsection{W-types} W-types have a standard object-wise inductive definition in
the presheaf model \cite{DBLP:journals/apal/MoerdijkP00}. Additionally, we need to give a logical relation
interpretation to W-types, as required for the soundness proof for staging. In
Section 5.2.\ of \cite{staged2ltt} a serialization map is used to define the relation
for natural numbers, however, there is no analogous map for W-types because we
cannot build finite terms from infinitely branching trees.

Instead, we give an inductive definition for the logical relation at
W-types. First, assuming $A : \wh{\Set}$ and $B : A \to \wh{\Set}$, we have the
usual constructor for the semantic W-type in the internal language of the
presheaf model, as $\msf{sup} : (a : A) \to (B\,a \to \msf{W}\,A\,B) \to
\msf{W}\,A\,B$.  Second, assuming $A : \Ty\,\Gamma$ and $B : \Ty\,(\Gamma \ext
A)$ in CFTT, we internalize the $\msf{sup}$ constructor which constructs a term
of a syntactic W-type from restricted terms:
\[ \mathbb{R}_{\msf{sup}} : (\alpha : \mathbb{R}\,A\,\gamma) \to \mathbb{R}\,(B \RA (\msf{W_{CFTT}}\,A\,B)[\msf{wk}])\,(\gamma,\,\alpha) \to \mathbb{R}\,(\msf{W_{CFTT}}\,A\,B)\,\gamma \]
Here, $\msf{W_{CFTT}}$ denotes the W-type former in CFTT, $B \RA
(\msf{W_{CFTT}}\,A\,B)[\msf{wk}]$ is a non-dependent function type in CFTT, and
$\msf{wk}$ is a weakening substitution.

Then, assuming $\gamma^{\approx} : \Gamma^{\approx}\,\gamma\,\gamma'$, we
inductively define a relation $\msf{W}^{\approx}$ between
$\msf{W}\,(\mbb{E}\,A\,\gamma)\,(\lam \alpha.\,\mbb{E}\,B\,(\gamma,\,\alpha))$
and $\mbb{R}\,(\msf{W_{CFTT}}\,A\,B)\,\gamma'$, which relates
$\msf{sup}\,\alpha\,f$ to $\mathbb{R}_{\msf{sup}}\,\alpha'\,f'$ if $\alpha$ is
related to $\alpha'$ at type $A$ and $f$ is related to $f'$ at type $B[\alpha']
\RA \msf{W_{CFTT}}\,A\,B$. The relation at the non-dependent function type $\blank{\RA}\blank$
is given as pointwise preservation of relations. With this, we can also give
relational interpretation to the syntactic $\msf{sup}$ rule in CFTT,
and we can use induction on $\msf{W}^{\approx}$ to interpret the syntactic
induction rule for W. Thus, we get soundness of staging with W-types.

\subsubsection{Identity type} Our treatment of the identity type is
influenced by the following: we would like to assume equations as axioms in
CFTT, which are true in the presheaf model (i.e.\ they are true at
staging-time), without blocking staging as a computation. In Section \ref{sec:concatmap-for-streams}
we use such an axiom to good effect. However, axioms do block computation in
CFTT up to definitional equality, which implies that the soundness of staging
fails. Our solution:
\begin{itemize}
\item Have a standard intensional identity type in CFTT.
\item Before staging, translate the intensional identity to extensional
      identity. This amounts to erasing all identity proofs and transports.
\item Perform staging after erasure of identity proofs, i.e.\ model extensional identity in the semantics.
\end{itemize}
In the presheaf model, we interpret the extensional identity in a standard way,
as strict equality of sections of presheaves, and soundness for extensional
identity becomes trivial to show. Hence, the soundness property for our overall
staging pipeline becomes the following: the staging output is definitionally
equal to the identity-erased version of the staging input.

@@

\item It seems promising to look for synergies between push streams, pull streams
     and monad transformers. It seems that pull streams could be generalized from
     $\Gen$ to different monads, in the representation of transitions and initial
     states. This would allow putting a $\Pull$ \emph{on the top} of a monad
     transformer stack. On the other hand, push streams form a proper monad, but
     they cannot be used to ``transform'' other monads, so they can be placed
     at the \emph{bottom} of a transformer stack.
\item We could try to lean more heavily on datatype-generic programming and
     generalize abstractions to arbitrary object types. For example, push and pull
     streams could be generalized to inductive and coinductive Church-codings of
     arbitrary value types.
\end{itemize}

>>>>

\usepackage{xcolor}
\usepackage{mathpartir}
\usepackage{todonotes}
\presetkeys{todonotes}{inline}{}
\usepackage{scalerel}
\usepackage{bm}
\usepackage{mathtools}
\usepackage{stmaryrd}
\usepackage[title]{appendix}

\newcommand{\mit}[1]{{\mathsf{#1}}}
\newcommand{\msf}[1]{{\mathsf{#1}}}
\newcommand{\mbf}[1]{{\mathbf{#1}}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\code}{\mathsf{code}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\wh}[1]{\widehat{#1}}
\newcommand{\mdo}{\mbf{do}\,}
\newcommand{\ind}{\hspace{1em}}
\newcommand{\bif}{\mbf{if}\,}
\newcommand{\bthen}{\mbf{then}\,}
\newcommand{\belse}{\mbf{else}\,}
\newcommand{\return}{\mbf{return}\,}
\newcommand{\pure}{\mbf{pure}\,}
\newcommand{\lam}{\lambda\,}
\newcommand{\data}{\mbf{data}\,}
\newcommand{\where}{\mbf{where}}
\newcommand{\M}{\msf{M}}
\newcommand{\letrec}{\mbf{letrec}\,}
\newcommand{\of}{\mbf{of}\,}
\newcommand{\go}{\mit{go}}
\newcommand{\add}{\mit{add}}
\newcommand{\letdef}{\mbf{let\,}}
\newcommand{\map}{\mit{map}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\Tyo}{\msf{Ty}_{\mbbo}}
\newcommand{\Tmo}{\msf{Tm}_{\mbbo}}
\newcommand{\whset}{\wh{\Set}}
\newcommand{\ev}{\mbb{E}}
\newcommand{\re}{\mbb{R}}

\newcommand{\mbbc}{\mbb{C}}
\newcommand{\mbbo}{\mbb{O}}

\newcommand{\vas}{\mathsf{as}}
\newcommand{\vbs}{\mathsf{bs}}
\newcommand{\vcs}{\mathsf{cs}}
\newcommand{\vxs}{\mathsf{xs}}
\newcommand{\vys}{\mathsf{ys}}
\newcommand{\vsp}{\mathsf{sp}}
\newcommand{\vma}{\mathsf{ma}}
\newcommand{\vga}{\mathsf{ga}}
\newcommand{\vm}{\mathsf{m}}
\newcommand{\vn}{\mathsf{n}}
\newcommand{\vk}{\mathsf{k}}
\newcommand{\vA}{\mathsf{A}}
\newcommand{\vB}{\mathsf{B}}
\newcommand{\vC}{\mathsf{C}}
\newcommand{\vS}{\mathsf{S}}
\newcommand{\vF}{\mathsf{F}}
\newcommand{\vR}{\mathsf{R}}
\newcommand{\vM}{\mathsf{M}}
\newcommand{\vmb}{\mathsf{mb}}
\newcommand{\mAs}{\mathsf{As}}
\newcommand{\va}{\mathsf{a}}
\newcommand{\vb}{\mathsf{b}}
\newcommand{\vc}{\mathsf{c}}
\newcommand{\vd}{\mathsf{d}}
\newcommand{\vx}{\mathsf{x}}
\newcommand{\vy}{\mathsf{y}}
\newcommand{\vz}{\mathsf{z}}
\newcommand{\vf}{\mathsf{f}}
\newcommand{\vfs}{\mathsf{fs}}
\newcommand{\vg}{\mathsf{g}}
\newcommand{\vh}{\mathsf{h}}
\newcommand{\vt}{\mathsf{t}}
\newcommand{\vs}{\mathsf{s}}
\newcommand{\vr}{\mathsf{r}}
\newcommand{\vu}{\mathsf{u}}
\newcommand{\vl}{\mathsf{l}}
\newcommand{\vns}{\mathsf{ns}}
\newcommand{\vW}{\mathsf{W}}
\newcommand{\vsup}{\mathsf{sup}}
\newcommand{\vid}{\mathsf{id}}
\newcommand{\whW}{\wh{\vW}}

@@

\newcommand{\Con}{\msf{Con}}
\newcommand{\Sub}{\msf{Sub}}
\newcommand{\Tm}{\msf{Tm}}
\newcommand{\Mor}{\msf{Tm}}

@@

\newcommand{\MTy}{\msf{MetaTy}}
\newcommand{\MTm}{\msf{MetaTm}}
\newcommand{\VTy}{\msf{ValTy}}
\newcommand{\Ty}{\msf{Ty}}
\newcommand{\CTy}{\msf{CompTy}}
\newcommand{\True}{\msf{True}}
\newcommand{\False}{\msf{False}}
\newcommand{\fst}{\msf{fst}}
\newcommand{\snd}{\msf{snd}}

@@

Modern functional programming supports many convenient abstractions. These often
come with significant runtime overheads. Sometimes the overheads are acceptable,
but in other cases compiler optimization is crucial. Monads in Haskell is an
example for the latter. Even the $\Reader$ monad, which is one of the simplest
in terms of implementation, yields large overheads when compiled without
optimizations. Consider the following:
\begin{alignat*}{3}
  &\vf :: \Int \to \Reader\,\Bool\,\Int \\
  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vx\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
\end{alignat*}
With optimizations enabled, GHC compiles this roughly to the code below:
\begin{alignat*}{3}
  &\vf :: \Int \to \Bool \to \Int \\
  &\vf = \lam \vx\,\vb.\, \bif \vb\, \bthen \vx + 10\, \belse \vx + 20
\end{alignat*}
Without optimizations we roughly get:
\begin{alignat*}{3}
  &\vf = \lam \vx.\,(\bind)\,\msf{MonadReaderDict}\,(\msf{ask}\,\msf{MonadReaderDict})\,(\lam \vb.\,\bif \vb\\
  &\ind \bthen\return\,\msf{MonadReaderDict}\,(\vx + 10)\\
  &\ind \belse\hspace{0.25em}\return\,\msf{MonadReaderDict}\,(\vx + 20))
\end{alignat*}
Here, $\msf{MonadReaderDict}$ is a runtime dictionary, containing the methods of
the $\Monad$ instance for $\Reader$, and $(\bind)\,\msf{MonadReaderDict}$ is the
projection of the binding function. Here, a runtime closure will be created for
the $\lam \vb.\,...$ function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$
will create additional dynamic closures.

The difference between optimized and unoptimized code is already large here, and
it gets even larger when we consider monad transformers or code that is
polymorphic over monads. In Haskell, such code is pervasive, even in fairly
basic programs which do not use fancy abstractions. Consider the $\msf{mapM}$
function from the Haskell Prelude:
\begin{alignat*}{3}
  & \msf{mapM} :: \Monad\,\vm \Rightarrow (\va \to \vm\,\vb) \to [\va] \to \vm\,[\vb]
\end{alignat*}
This is a third-order and second-rank polymorphic function in disguise, because
its monad dictionary argument contains the polymorphic second-order method
$(\bind)$.  Compiling $\msf{mapM}$ efficiently relies on inlining the instance
dictionary, then inlining the methods contained there, and also inlining the
functions that the higher-order binding is applied to.

@@

\subsection{Staged Semantics of CFTT}\label{sec:semantics-of-staging}

We give an overview of the unstaging algorithm together with its correctness.
We reuse the concepts and most of the results from \cite{staged2ltt}. Our main
task is to extend ibid.\, with identity types and inductive types in the
metalanguage, since we use them in this paper. For inductive types, it
is the most sensible to develop W-types, since all inductive families can be
faithfully derived from them in our setting \cite{whynotw}.

This however requires a substantial amount of technical background from
\cite{staged2ltt}, so we relegate it to Appendix \ref{appendix:A}. Here we only
review the basic notions.
\begin{itemize}
\item The input of unstaging is a CFTT term with an object-level type, which
      only depends on object-level free variables. The output is a term in the
      object theory. In the other direction, there is an evident embedding of
      object-theoretic terms as CFTT terms.
\item \emph{Soundness} means that unstaging followed by embedding is the
      identity map, up to conversion. In other words, unstaging respects conversion
      of CFTT terms.
\item \emph{Stability} means that embedding followed by unstaging is the
      identity map. In other words, unstaging has no action on terms which are
      already purely object-level (i.e.\ contain no splices).
\end{itemize}
In \cite{staged2ltt}, there is an extra property, called \emph{strictness},
which means that unstaging does not perform $\beta\eta$-conversion in the
object theory, but this holds trivially in our case since we do not have such
conversion rules.

However, for CFTT we slightly depart from the above notion of soundness, for the
following reason. We would like to assume certain propositional identities as
\emph{axioms} in the metalanguage, without blocking unstaging as a
computation. We use such an axiom to good effect in Section
\ref{sec:concatmap-for-streams}.

Now, axioms clearly block computation, so they are incompatible with the
mentioned notion of soundness. But we are only interested in equations which
hold definitionally during unstaging. Hence, we can do the following: first, we
erase all identity proofs and their transports from CFTT programs, then we
proceed with unstaging. The erasure of identity proofs can be defined as a
syntactic translation \cite{next700} from CFTT into CFTT with an extensional
identity type. Thus, what we get is that unstaging is sound only after the
erasure of identity proofs.

%% \subsubsection{Identity type} Our treatment of the identity type is
%% influenced by the following: we would like to assume equations as axioms in
%% CFTT, which are true in the presheaf model (i.e.\ they are true at
%% staging-time), without blocking staging as a computation. In Section \ref{sec:concatmap-for-streams}
%% we use such an axiom to good effect. However, axioms do block computation in
%% CFTT up to definitional equality, which implies that the soundness of staging
%% fails. Our solution:
%% \begin{itemize}
%% \item Have a standard intensional identity type in CFTT.
%% \item Before staging, translate the intensional identity to extensional
%%       identity. This amounts to erasing all identity proofs and transports.
%% \item Perform staging after erasure of identity proofs, i.e.\ model extensional identity in the semantics.
%% \end{itemize}
%% In the presheaf model, we interpret the extensional identity in a standard way,
%% as strict equality of sections of presheaves, and soundness for extensional
%% identity becomes trivial to show. Hence, the soundness property for our overall
%% staging pipeline becomes the following: the staging output is definitionally
%% equal to the identity-erased version of the staging input.

%% In the
%% current work we use identity types and some inductive types and the
%% metalanguage, which are not covered in \cite{staged2ltt}. Hence, our main task

%% We give an overview of the semantics of CFTT. This yields a algorithm for
%% evaluating
%% extracting object programs from CFTT

%% We give a short summary of the semantics of staging here. We reuse most of the
%% semantics from \cite{staged2ltt} and only mention additions and differences. The
%% reader may refer to ibid.\ for technical details.

%% Staging is given by evaluation of CFTT into a presheaf model, where the base
%% category is given as follows: objects are typing contexts of the object language
%% and morphisms are parallel substitutions. The equality of morphisms is given by
%% strict syntactic equality of object terms. $\Ty$ is interpreted as the presheaf
%% which is constantly the set of object-theoretic types. Soundness of staging is
%% proven by a logical relation model, internally to the mentioned presheaf
%% model. Soundness means that the output of staging, viewed as a CFTT term, is
%% definitionally equal to the input.

%% The main addition in CFTT is the assumption of all inductive families in the
%% meta-level. However, based on the results in \cite{whynotw}, it is enough to
%% extend the meta-level syntax in \cite{staged2ltt} with W-types and identity types to
%% derive all inductive families. We consider the semantics of these in the following.

%% \subsubsection{W-types} W-types have a standard object-wise inductive definition in
%% the presheaf model \cite{DBLP:journals/apal/MoerdijkP00}. Additionally, we need to give a logical relation
%% interpretation to W-types, as required for the soundness proof for staging. In
%% Section 5.2.\ of \cite{staged2ltt} a serialization map is used to define the relation
%% for natural numbers, however, there is no analogous map for W-types because we
%% cannot build finite terms from infinitely branching trees.

%% Instead, we give an inductive definition for the logical relation at
%% W-types. First, assuming $A : \wh{\Set}$ and $B : A \to \wh{\Set}$, we have the
%% usual constructor for the semantic W-type in the internal language of the
%% presheaf model, as $\msf{sup} : (a : A) \to (B\,a \to \msf{W}\,A\,B) \to
%% \msf{W}\,A\,B$.  Second, assuming $A : \Ty\,\Gamma$ and $B : \Ty\,(\Gamma \ext
%% A)$ in CFTT, we internalize the $\msf{sup}$ constructor which constructs a term
%% of a syntactic W-type from restricted terms:
%% \[ \mathbb{R}_{\msf{sup}} : (\alpha : \mathbb{R}\,A\,\gamma) \to \mathbb{R}\,(B \RA (\msf{W_{CFTT}}\,A\,B)[\msf{wk}])\,(\gamma,\,\alpha) \to \mathbb{R}\,(\msf{W_{CFTT}}\,A\,B)\,\gamma \]
%% Here, $\msf{W_{CFTT}}$ denotes the W-type former in CFTT, $B \RA
%% (\msf{W_{CFTT}}\,A\,B)[\msf{wk}]$ is a non-dependent function type in CFTT, and
%% $\msf{wk}$ is a weakening substitution.

%% Then, assuming $\gamma^{\approx} : \Gamma^{\approx}\,\gamma\,\gamma'$, we
%% inductively define a relation $\msf{W}^{\approx}$ between
%% $\msf{W}\,(\mbb{E}\,A\,\gamma)\,(\lam \alpha.\,\mbb{E}\,B\,(\gamma,\,\alpha))$
%% and $\mbb{R}\,(\msf{W_{CFTT}}\,A\,B)\,\gamma'$, which relates
%% $\msf{sup}\,\alpha\,f$ to $\mathbb{R}_{\msf{sup}}\,\alpha'\,f'$ if $\alpha$ is
%% related to $\alpha'$ at type $A$ and $f$ is related to $f'$ at type $B[\alpha']
%% \RA \msf{W_{CFTT}}\,A\,B$. The relation at the non-dependent function type $\blank{\RA}\blank$
%% is given as pointwise preservation of relations. With this, we can also give
%% relational interpretation to the syntactic $\msf{sup}$ rule in CFTT,
%% and we can use induction on $\msf{W}^{\approx}$ to interpret the syntactic
%% induction rule for W. Thus, we get soundness of staging with W-types.

%% \subsubsection{Identity type} Our treatment of the identity type is
%% influenced by the following: we would like to assume equations as axioms in
%% CFTT, which are true in the presheaf model (i.e.\ they are true at
%% staging-time), without blocking staging as a computation. In Section \ref{sec:concatmap-for-streams}
%% we use such an axiom to good effect. However, axioms do block computation in
%% CFTT up to definitional equality, which implies that the soundness of staging
%% fails. Our solution:
%% \begin{itemize}
%% \item Have a standard intensional identity type in CFTT.
%% \item Before staging, translate the intensional identity to extensional
%%       identity. This amounts to erasing all identity proofs and transports.
%% \item Perform staging after erasure of identity proofs, i.e.\ model extensional identity in the semantics.
%% \end{itemize}
%% In the presheaf model, we interpret the extensional identity in a standard way,
%% as strict equality of sections of presheaves, and soundness for extensional
%% identity becomes trivial to show. Hence, the soundness property for our overall
%% staging pipeline becomes the following: the staging output is definitionally
%% equal to the identity-erased version of the staging input.

@@

\item It seems promising to look for synergies between push streams, pull streams
     and monad transformers. It seems that pull streams could be generalized from
     $\Gen$ to different monads, in the representation of transitions and initial
     states. This would allow putting a $\Pull$ \emph{on the top} of a monad
     transformer stack. On the other hand, push streams form a proper monad, but
     they cannot be used to ``transform'' other monads, so they can be placed
     at the \emph{bottom} of a transformer stack.
\item We could try to lean more heavily on datatype-generic programming and
     generalize abstractions to arbitrary object types. For example, push and pull
     streams could be generalized to inductive and coinductive Church-codings of
     arbitrary value types.
\end{itemize}
\newpage

\appendices
\section{Semantics of CFTT}\label{appendix:A}

In this appendix we describe the semantics of CFTT, from which the unstaging
algorithm is obtained, and also describe the soundness and completeness of the
algorithm. We borrow the setup from \cite{staged2ltt}; we use terminology and
definitions from there by default, and focus on differences and extensions.

\subsection{Models and Syntax of CFTT}

We give an algebraic \cite{TODO} specification for the models of CFTT and define
the syntax of CFTT to be the initial model.

\begin{definition}[] A model of CFTT consists of the following.

\begin{itemize}
\item A category with a terminal object; in the syntax this is the category
  of contexts and parallel substitutions. We use $\Con : \Set$ for contexts,
  $\Mor : \Con \to \Con \to \Set$ for morphisms and $\emptycon : \Con$ for the
  terminal object (which is the empty context in the syntax).
\item A family structure in the sense of category-with-families \cite{TODO},
  specifying metatypes and terms. This structure is additionally indexed by
  universe levels, so we have $\MTy_i : \Con \to \Set$ and $\msf{MetaTm}_i :
  (\Gamma : \Con) \to \MTy_i\,\Gamma \to \Set$ for $i \in \mbb{N}$. We support
  basic type formers, identity types and W-types. Additionally, we have
  Coquand-style universes, i.e.\ $\U_i : \MTy_{i+1}\,\Gamma$ such that there is
  a definitional isomorphism $\msf{MetaTm}_{i+1}\,\Gamma\,\U_i \simeq \MTy_i\,\Gamma$.
  We use $\El$ and $\code$ to denote the components of this isomorphism.
\item
  For object-level types, we have $\Ty : \MTy_0\,\Gamma$, $\VTy :
  \MTy_0\,\Gamma$ and $\CTy : \MTy_0\,\Gamma$, together with all object-level
  type formers and inclusion operations $\msf{V} : \MTm\,\Gamma\,\VTy \to \MTm\,\Gamma\,\Ty$
  and $\msf{C} : \MTm\,\Gamma\,\CTy \to \MTm\,\Gamma\,\Ty$.
\item For object-level terms, we have $\Tm : \MTm\,\Gamma\,\Ty \to \Set$, closed under
  all object-level term formers. Here, $\MTm\,\blank\,\Ty$ and $\Tm$ also form a
  family structure, so that we can substitute object terms and extend the context
  with object term variables.
\item Staging operations: we have $\Lift : \MTm\,\Gamma\,\Ty \to \MTy\,\Gamma$ together
  with definitional isomorphisms $\Tm\,\Gamma\,A \simeq \MTm\,\Gamma\,(\Lift A)$.
  We write $\ql\!\blank\!\qr$ and $\spl\!\blank$ for the components of the isomorphism.
\end{itemize}
\end{definition}
This presentation is a bit more verbose than the ``surface'' syntax used in the
paper. First, the meta level uses explicit coding and decoding operations for
universes, while the surface syntax assumes Russell-style universes. Second, the
object level has $\VTy$ and $\CTy$ as Tarski-style universes with explicit
inclusion into $\Ty$, while the surface syntax used implicit coercions.

Also, this is not the only possible presentation. Here, we follow the style of
the informal syntax by having a separate \emph{sort} for object-level terms,
together with explicit quoting and splicing, but we specify object-level types
as certain meta-level terms. This a hybrid solution of two styles:
\begin{itemize}
\item One would have separate sorts for $\Ty$, $\CTy$ and $\VTy$, and would have
  separate staging operations for all of these sorts. This follows the ``classic''
  2LTT presentation \cite{TODO}, where the two levels are specified independently
  and level-mixing is only possible through the staging operations.
\item The other one would only have metatypes and terms, no lifting, quoting or
  splicing, and would use metafunctions to represent object-level binders. This
  is the HOAS (``higher-order abstract syntax'') or ``logical framework''
  flavor. Our Agda implementation actually follows this style.
\end{itemize}
In the context of CFTT, the choice between HOAS and 2LTT styles is a matter of
taste, and the two are inter-derivable.

\subsection{Presheaf Model}

Unstaging is defined as evaluation of the syntax of CFTT in a particular
presheaf model.
\begin{itemize}
  \item The \emph{object theory} is the simple type theory supporting only value
    types and computation types as described in Section
    \ref{sec:the-object-level} and also in the Agda
    formalization. Object-theoretic terms are not quotiented by
    $\beta\eta$-conversion. Substitution is a recursively defined operation on
    terms.
  \item $\mbbo$ is the category whose objects are contexts of the object theory,
    and morphisms are parallel substitutions, as lists of terms. With this,
    object-theoretic types form a constant presheaf which we denote
    as $\Tyo : \mbbo^{\msf{op}} \to \Set$, and terms form a dependent presheaf over $\Tyo$ which we write as
    $\Tmo$.
  \item
  In the presheaf model of CFTT, every typing context (and also every closed
  type) is modeled as a presheaf over $\mbbo$. The interpretation of meta-level
  type formers is standard, following Hofmann \cite{Hofmann97syntaxand} and
  Huber \cite{huber-thesis}.
\item $\Ty : \MTy_0\,\Gamma$ is the displayed presheaf which is constantly
  $\Tyo$. $\VTy$ and $\CTy$ are modeled analogously.
\item $\Lift A : \MTy_0\,\Gamma$ is modeled as follows. We have that $A :
  \MTy\,\Gamma\,\Ty$, but since $\Ty$ is defined as constantly $\Tyo$, it
  follows that $A$ is a natural transformation from $\Gamma$ to $\Tyo$. Hence,
  we define $\Lift A : \MTy_0\,\Gamma$ to be $\Tmo[A]$, where $[\blank]$ is
  substitution in presheaves, i.e.\ the reindexing of a displayed presheaf by a
  natural transformation.
\item
  $\Tm\,\Gamma\,A$ is simply defined to be $\MTy\,\Gamma\,(\Lift A)$ in the model, which
  means that quotation and splicing are both identity maps.
\end{itemize}
Now, we can take a closed term $t : \Tm\,\emptycon\,A$ in CFTT and evaluate it
in the presheaf model to obtain an object-theoretic term whose type is the
result of evaluating $A$.

\subsection{Soundness for W-types}
So far we have stayed close to \cite{staged2ltt}, with only cosmetic
changes. Supporting W-types, as a way to get inductive families \cite{TODO}, is
the biggest extension here. We extend Section 5 of \cite{staged2ltt} here.  In
the syntax of CFTT, we have
\begin{alignat*}{3}
  &\vW   &&: (A : \MTy_i\,\Gamma) \to \MTy_j\,(\Gamma,\,A) \to \MTy_{\msf{max}(i,j)}\,\Gamma\\
  &\vsup &&: (a : \MTm\,\Gamma\,A) \to \MTm\,(\Gamma,B[\vid,a])\,(\vW\,A\,B) \to \MTm\,\Gamma\,(\vW\,A\,B),
\end{alignat*}
where $\vid$ is the identity substitution from $\Gamma$ to $\Gamma$. We also have the eliminator for $\vW$
whose specification we omit here.

Internally to $\whset$, i.e.\ internally to presheaves over $\mbbo$, we also
have the standard W-type, defined pointwise as the W-type in sets:
\begin{alignat*}{3}
  & \wh{\vW}   &&: (A : \whset) \to (A \to \whset) \to \whset\\
  & \wh{\vsup} &&: (a : A) \to (B\,a \to \wh{\vW}\,A\,B) \to \wh{\vW}\,A\,B
\end{alignat*}
In the presheaf model we can evidently interpret $\vW$ using $\whW$.

%% The \emph{restriction} of this

%% \subsubsection{Stability} Stability means that if we take any object-theoretic term,
%% embed it as a CFTT term, then unstage it
